PACKER: An Exemplar Model of Category Generation

### 2017-01-30 Manuscripts/cogsci-2017/paper.tex
slight edits

Generating new concepts is an intriguing yet understudied topic in cognitive science. In this paper, we present a novel exemplar model of category generation: PACKER (\textit{Producing Alike and Contrasting Knowledge using Exemplar Representations}). PACKER's core design assumptions are (1) categories are represented as exemplars in a multidimensional psychological space, (2) generated items should be similar to exemplars of the same category, and (3) generated categories should be dissimilar to existing categories. A behavioral study and formal simulations broadly support PACKER's predictions.

@@

In a behavioral study and subsequent formal modeling, we found broad support for the PACKER model. Participants in our study more frequently generated items that are distant from members of contrast categories, and they tended to generate categories with more within-class than between-class similarity. Likewise, we found that the \textit{location} of contrast categories shapes generation by imposing constraints on the areas of space that remain available for a new category. Existing models \citep[see][]{jern2013probabilistic}, possessing no mechanisms for category-contrast, do not account for these effects. Our formal simulations reveal that PACKER provides a strong account of the phenomena observed behaviorally.

The PACKER model is, in general, highly expressive in its performance. Under different parameter settings it is capable of generating tightly clustered or highly distributed categories, and adjusting the distribution of categories along each feature. Future work will focus on exploring individual differences in generation, and whether PACKER can explain previous results in the field \citep{ward1994structured,jern2013probabilistic}.

>>>>

Generating new concepts is an intriguing yet understudied topic in cognitive science. In this paper, we present a novel exemplar model of category generation: PACKER (\textit{Producing Alike and Contrasting Knowledge using Exemplar Representations}). PACKER's core design assumptions are (1) categories are represented as exemplars in a multidimensional psychological space, (2) generated items should be similar to exemplars of the same category, and (3) generated categories should be dissimilar to existing categories. A behavioral study reveals strong effects of contrast- and target-class similarity, which are directly predicted by the PACKER model but are not explained by existing formal approaches.

@@

In a behavioral study and subsequent formal modeling, we found broad support for the PACKER model. Participants in our study more frequently generated items that are distant from members of contrast categories, and they tended to generate categories with more within-class than between-class similarity. Likewise, we found that the \textit{location} of contrast categories (as opposed to their structure) shapes generation by imposing constraints on the areas of space that remain available for a new category. Formal simulations reveal that existing models \citep[see][]{jern2013probabilistic}, making no assumptions about category-contrast, do not account for these effects.

The PACKER model is, in general, highly expressive in its performance. Under different parameter settings it is capable of generating tightly clustered or highly distributed categories, and adjusting the distribution of categories along each feature. Future work will focus on exploring the broad degree of individual differences we observed in generation, and whether PACKER can explain previous results in the field \citep{ward1994structured,jern2013probabilistic}.
